---
title: Generating selectively reported results in meta-analysis
date: '2026-04-07'
categories:
- meta-analysis
- publication bias
- selective reporting
code-fold: true
code-tools: true
toc: true
bibliography: "../selection-references.bib"
csl: "../apa.csl"
link-citations: true
draft: true
---

For one of her dissertation studies, my advisee [Paulina Grekov](/people/Paulina-Grekov/) is looking at factors that influence the performance of publication bias adjustment methods in meta-analysis (e.g., methods like Trim-and-Fill, PET-PEESE, and the three-parameter selection model). 
Of course, there is already a large literature about these methods, including many simulation studies that have looked at adjustment methods under a range of different conditions [e.g.,@bramley2021examining; @carter2019correcting; @mcshane2016adjusting; @stanley2014metaregression;  @vanAert2026correcting, to name but a few].
Past simulations have looked at a variety of adjustment methods under a wide variety of different data-generating processes, varying in the metric of the simulated effect size estimates, the distribution of primary study sample sizes, the form and strength of selective reporting, and other factors. 
@hong2021using reviewed many of these past studies, demonstrating that differences in the conditions examined (what they call the _simulation environment_) lead to discrepant findings about the performance of some methods. 
How well the bias-adjustment methods work (and which method or methods should be used in practice) seems to depend on how the data were generated, such as the degree of heterogeneity, how large the simulated studies are, and how severe the selection bias is.

One thing that makes it challenging to compare the findings from these past studies is that they use procedures for simulating meta-analytic datasets. 
One thing that Paulina is interested in is how much the procedure itself matters. 
Many past studies simulated datasets based on some more-or-less simple form of [step-function selection model](/posts/step-function-selection-models), such as the so-called three-parameter selection model (3PSM) that has a single step at $\alpha = .025$.
But there are different ways to actually program this, and it's a little tricky to line them up. 
In this post, I'll look at two such methods: one used by @mcshane2016adjusting and one by @stanley2014metaregression.

## Stochastic selection

The 3PSM posits that effect size estimates are observed based on a two-step process.
First, estimates are generated (but potentially not observed) based on a random effects model, in which
$$
T_i^* \sim N\left(\mu, \ \tau^2 + {\sigma_i^*}^2\right),
$$
where I use the asterisks on $T_i^*$ and $\sigma_i^*$ to indicate that the estimates are generated but not necessarily observed.
Second, the effects are censored according to whether their one-sided p-values are above or below $\alpha_1 = .025$.
If $p_i^* = 1 - \Phi^{-1}(T_i^* / \sigma_i^*)$ is less than $\alpha_1$, then $\left(T_i^*, \sigma_i^*\right)$ is observed; if $p_i^* \geq \alpha_1$, then $\left(T_i^*, \sigma_i^*\right)$ is observed with probability $\lambda_1$ and censored with probability $1 - \lambda_1$.
If we let $a_i = I(p_i < \alpha_1)$ be an indicator for whether effect size estimate $i$ is affirmative (positive and statistically significant), then
$$
\Pr(T_i^* \text{ is observed} | T_i^*, \sigma_i^*) = a_i + \lambda_1 (1 - a_i).
$$

These steps lead to a distributional model for the _observed_ effect size estimates that is piece-wise normal.
Letting $\eta_i = \sqrt{\tau^2 + \sigma_i^2}$ and $c_{1i} = \left[\sigma_i \Phi^{-1}(1 - \alpha_1) - \mu\right] / \eta_i$, the density of an observed effect size estimate is given by
$$
\Pr(T_i = t | \sigma_i) = \frac{a_i + \lambda_1 (1 - a_i)}{1 - (1 - \lambda_1) \Phi(c_{1i})} \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right).
$$ {#eq-3PSM-density}

@mcshane2016adjusting simulated from the 3PSM using a stochastic selection process that, in my view, aligns very closely with the formulation of the model.
Their approach is amounts to a simple rejection sampler:
they simulate data based on each step of the process and repeat until they end up with a dataset that includes a total of $K$ observed effect size estimates. 
Under their approach, the expected value of an observed effect size estimate is 
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \frac{(1 - \lambda_1) \phi(c_{1i})}{1 - (1 - \lambda_1)\Phi(c_{1i})}.
$$ {#eq-3PSM-expectation}
In simulating a meta-analytic dataset, they start by drawing a primary study sample size, which means that $\sigma_i$ is itself a random quantity.
Thus, the unconditional expectation of $T_i$ will depend on the distribution of $\sigma_i$ used to generate the data.

## Mixture selection

@stanley2014metaregression and @stanley2017limitations reported simulations based on a subtly different appoach to implementing the selective reporting process.
Rather than treating the selection process as stochastic, given the p-value of the generated effect size estimate, they conceptualized the process as a mixture of two types of studies (or two types of researchers).
One type of researcher, comprising $\pi_1$ of the population, reports all results regardless of whether they are affirmative.
The other type of researcher, comprising the remaining $1 - \pi_1$ of the population, only reports results that are positive and statistically significant.
In the @stanley2014metaregression simulations, every meta-analytic dataset consists of $k \times \pi_1$ effect size estimates generated by the first type of researcher and $k \times (1 - \pi_1)$ effect size estimates generated by the second type.
Both types of researchers draw primary study sample sizes from the same distribution (i.e., researcher type is independent of $\sigma_i^*$).
The observed effect size estimates therefore follow a mixture distribution in which
$$
\begin{aligned}
\Pr(T_i = t | \sigma_i) &= \pi_1 \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right) + (1 - \pi_1) \times \frac{a_i \phi\left(\frac{t - \mu}{\eta_i}\right)}{\eta_i\left[1 - \Phi(c_{1i}) \right]} \\
&= \left[\pi_1+ (1 - \pi_1) \times \frac{a_i}{\left[1 - \Phi(c_{1i}) \right]} \right] \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right),
\end{aligned}
$$ {#eq-mixture-density}
with expectation
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \frac{(1 - \pi_1) \phi(c_{1i})}{1 - \Phi(c_{1i})}.
$$ {#eq-mixture-expectation}
The question is then how the mixing proportion $\pi_1$ relates to the selection parameter $\lambda_1$.

If we wanted to directly equate the distributions for a given standard error $\sigma_i$, we could set the mixing proportion to 
$$
\pi_{1i} = \frac{\lambda_1}{1 - (1 - \lambda_1)\Phi(c_{1i})}.
$$
The difficulty here is that the mixing proportion is a function of the standard error for effect size $i$, so studies with different standard errors would need to have different mixing proportions. 
That's not how @stanley2014metaregression implement their data-generating process, so it doesn't really work as a way to translate between the selection parameter and the mixing proportion.
However, we could try to equate the unconditional expectations, so that $\mathbb{E}(T_i)$ is held constant.
Doing so leads to
$$
\left(1 - \pi_1\right) \mathbb{E}\left[\frac{\eta_i \phi(c_{1i})}{1 - \Phi(c_{1i})}\right] = \left(1 - \lambda_1\right) \mathbb{E}\left[\frac{\eta_i \phi(c_{1i})}{1 - (1 - \lambda_1)\Phi(c_{1i})}\right],
$$ {#eq-equating-expectations}
which implies that
$$
\pi_1 = \lambda_1 \times \frac{\mathbb{E}\left[\frac{\eta_i \phi(c_{1i})}{\left[1 - \Phi(c_{1i})\right]\left[1 - (1 - \lambda_1)\Phi(c_{1i})\right]}\right]}{\mathbb{E}\left[\frac{\eta_i \phi(c_{1i})}{1 - \Phi(c_{1i})}\right]}.
$$ {#eq-pi1-marginally-equated}
Note that $c_{1i}$ is a function of $\mu$, $\tau$, and $\sigma_i$, and $\eta_i$ depends on $\tau$ and $\sigma_i$. Thus, the translation from $\lambda_1$ to $\pi$ will generally depend on _all_ the parameters of the model as well as on the distribution of $\sigma_i$'s.[^simpler-approx]

[^simpler-approx]: A slightly simpler approximate equating would be to take
$$
\pi_{1}' = \frac{\lambda_1}{1 - (1 - \lambda_1)\mathbb{E}\left[\Phi(c_{1i})\right]}.
$$

Computing $\pi_1$ will require some way of approximating the expectations in @eq-pi1-marginally-equated.
If $\sigma_i$ only takes on a finite set of distinct values, the expectation could be computed directly.
Otherwise, stochastic approximation could be used. 
For instance, suppose that $4 / \sigma_i^2 \sim Poisson(30)$, so we can simulate a large set of $\sigma_i$'s by taking
```{r}
#| code-fold: show

sigma_i <- 2 / sqrt(rpois(10000, lambda = 30))
```
Given values for the other model parameters, we compute the expectations by averaging over all the $\sigma_i$'s. Here is a function that computes $\pi_1$ given $\mu$, $\tau$, $\lambda_1$, $\alpha_1$, and a vector of $\sigma_i$ values:
```{r}
#| code-fold: show

mixing_prop <- function(
  mu, tau, lambda_1, sigma_i, 
  alpha_1 = .025
) {
  eta_i <- sqrt(tau^2 + sigma_i^2)
  c_1i <- (sigma_i * qnorm(1 - alpha_1) - mu) / eta_i
  d_1i <- dnorm(c_1i)
  P_1i <- pnorm(c_1i)
  num_i  <- eta_i * d_1i / ((1 - P_1i) * (1 - (1 - lambda_1) * P_1i))
  denom_i <- eta_i * d_1i / (1 - P_1i)
  pi_1 <- lambda_1 * mean(num_i) / mean(denom_i)
  return(pi_1)
}

mixing_prop(mu = 0.2, tau = 0.1, lambda_1 = 0.2, sigma_i = sigma_i)

```

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: 100%
#| fig-column: body
#| fig-cap: "Mixing proportion $\\pi_1$ as a function of selection parameter $\\lambda_1$ for selected values of $\\mu$ and $\\tau$."
#| lightbox: true
#| label: fig-mixing-prop

library(tidyverse)
param_grid <- 
  expand_grid(
    mu = seq(0.0, 0.6, 0.2),
    tau = seq(0.0, 0.4, 0.1),
    lambda_1 = seq(0, 1, 0.02)
  ) %>%
  mutate(
    mu_f = formatC(mu, format = "f", digits = 1) |> 
      fct() |> 
      fct_relabel(\(x) paste("mu ==", x)),
    tau_f = formatC(tau, format = "f", digits = 1) |> fct(),
    pi_1 = pmap_dbl(
      .l = list(mu = mu, tau = tau, lambda_1 = lambda_1),
      .f = mixing_prop, 
      sigma_i = sigma_i
    )
  )

ggplot(param_grid) + 
  aes(lambda_1, pi_1, color = tau_f) + 
  geom_line() + 
  facet_wrap(~ mu_f, labeller = "label_parsed") + 
  theme_minimal() + 
  theme(legend.position = "top") + 
  labs(
    x = expression(lambda[1]),
    y = expression(pi[1]),
    color = expression(tau)
  )
```

@fig-mixing-prop shows how the mixing proportion changes as a function of the selection parameter $\lambda_1$ for several different values of the average effect size $\mu$ and between-study heterogeneity $\tau$, using the same set of $\sigma_i$'s simulated above.
It is a fairly complicated function.
For smaller average effects, the degree of heterogeneity has relatively more influence on the shape of the function, whereas for larger average effects, the degree of heterogeneity hardly matters.
And everything depnds on the $\sigma_i$ distribution;  changing it would shift all of the curves in the picture. 

## More steps

