---
title: Generating selectively reported results in meta-analysis
date: '2026-04-07'
categories:
- meta-analysis
- publication bias
- selective reporting
code-fold: true
draft: true
code-tools: true
toc: true
bibliography: "../selection-references.bib"
csl: "../apa.csl"
link-citations: true
---

For one of her dissertation studies, my advisee [Paulina Grekov](/people/Paulina-Grekov/) is looking at factors that influence the performance of publication bias adjustment methods in meta-analysis (e.g., methods like Trim-and-Fill, PET-PEESE, and the three-parameter selection model). 
Of course, there is already a large literature about these methods, including many simulation studies that have looked at adjustment methods under a range of different conditions [e.g.,@bramley2021examining; @carter2019correcting; @mcshane2016adjusting; @stanley2014metaregression;  @vanAert2026correcting, to name but a few].
Past simulations have looked at a variety of adjustment methods under a wide variety of different data-generating processes, varying in the metric of the simulated effect size estimates, the distribution of primary study sample sizes, the form and strength of selective reporting, and other factors. 
@hong2021using reviewed many of these past studies, demonstrating that differences in the conditions examined (what they call the _simulation environment_) lead to discrepant findings about the performance of some methods. 
How well the bias-adjustment methods work (and which method or methods should be used in practice) seems to depend on how the data were generated, such as the degree of heterogeneity, how large the simulated studies are, and how severe the selection bias is.

One thing that makes it challenging to compare findings across these past studies is that they use different procedures for simulating meta-analytic datasets. 
Many past studies simulated datasets based on some more-or-less simple form of [step-function selection model](/posts/step-function-selection-models), such as the so-called three-parameter selection model (3PSM) that has a single step at $\alpha = .025$.
But there are different ways to actually program this, and it's a little tricky to line them up. 
In this post, I'll look at two such methods: one used by @mcshane2016adjusting and one by @stanley2014metaregression.

## Stochastic selection

The 3PSM posits that effect size estimates are observed based on a two-step process.
First, estimates are generated (but potentially not observed) based on a random effects model, in which
$$
\left(T_i^* | \sigma_i^*\right) \sim N\left(\mu, \ \tau^2 + {\sigma_i^*}^2\right),
$$ {#eq-random-effects-model}
where I use the asterisks on $T_i^*$ and $\sigma_i^*$ to indicate that the estimates are generated but not necessarily observed.
Second, the effects are censored according to whether their one-sided p-values are above or below $\alpha_1 = .025$.
Let me denote the one-sided p-value as $p_i^* = 1 - \Phi^{-1}(T_i^* / \sigma_i^*)$, let $a_i^* = I(p_i^* < \alpha_1)$ be an indicator for whether effect size estimate $i$ is affirmative (positive and statistically significant), and let $O_i$ be a binary indicator for whether $\left(T_i^*, \sigma_i^*\right)$ is observed.
The 3PSM posits that 
$$
\Pr(O_i = 1 | T_i^*, \sigma_i^*) = a_i^* + \lambda_1 (1 - a_i^*),
$$ {#eq-selection-process-3PSM}
so if $p_i^* \geq \alpha_1$, then $\left(T_i^*, \sigma_i^*\right)$ is observed with probability $\lambda_1$ and censored with probability $1 - \lambda_1$.

As described in [my earlier post](/posts/step-function-selection-models), these assumptions lead to a distributional model for the _observed_ effect size estimates that is piece-wise normal.
Letting $\eta_i = \sqrt{\tau^2 + \sigma_i^2}$ and $c_{1i} = \left[\sigma_i \Phi^{-1}(1 - \alpha_1) - \mu\right] / \eta_i$, the density of an observed effect size estimate is given by
$$
\Pr(T_i = t | \sigma_i) = \frac{a_i + \lambda_1 (1 - a_i)}{1 - (1 - \lambda_1) \Phi(c_{1i})} \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right).
$$ {#eq-3PSM-density}

@mcshane2016adjusting simulated from the 3PSM using a stochastic selection process that, in my view, aligns very closely with the formulation of the model.
Their approach is as follows:

1. Generate $\sigma_i^*$ by sampling from some distribution $f_\sigma$
2. Generate $T_i^*$ according to @eq-random-effects-model.
3. Simulate $O_i$ according to @eq-selection-process-3PSM.
4. If $O_i = 1$, then keep $\left(T_i^*, \sigma_i^*\right)$. Otherwise, start over.

They repeat this until they end up with a dataset that includes a total of $K$ observed effect size estimates. 

To compare their approach to other approaches, I would like to look at the unconditional expectation of $T_i$, which requires a bit more work because $\sigma_i$ is itself a random quantity.
Under the 3PSM, the conditional expected value of the observed effect size estimate is 
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \frac{(1 - \lambda_1) \phi(c_{1i})}{1 - (1 - \lambda_1)\Phi(c_{1i})}.
$$ {#eq-3PSM-expectation}
The unconditional expectation of $T_i$ will depend on the distribution of $\sigma_i$ used to generate the data---specifically, the distribution of the _observed_ $\sigma_i$, which depends on the probability that $\left(T_i^*, \sigma_i^*\right)$ is observed.
Conditional on $\sigma_i$ (but not on $T_i$), 
$$
\Pr(O_i = 1 | \sigma_i^*) = 1 - (1 - \lambda_1) \Phi(c_{1i}).
$$ {#eq-observation-prob-3PSM}

Also, it's perhaps a bit weird to worry about the unconditional expectation, considering that meta-analysts will typically focus on weighted averages of the effect size estimates, as in a random effects model. 
In practice, the weights will depend on estimation of the between-study heterogeneity $\tau$, but we can gloss over this by considering random effects weights based on the true value of $\tau$. 
Because this quantity won't actually ever be known in real life, I'll call this the oracle-weighted expectation.
Letting $\mathbb{E}$ be the expectation with respect to the distribution of observed $\sigma_i$'s and taking $\omega = \mathbb{E}(\eta_i^{-2})$, the unconditional oracle-weighted expectation of $T_i$ is then 
$$
\mathbb{E} \left[\frac{\mathbb{E}(T_i | \sigma_i)}{\omega \eta_i^2} \right] = \mu + (1 - \lambda_1) \times \mathbb{E}\left[\frac{\phi(c_{1i})}{\omega \eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i})\right]} \right].
$$

## Mixture selection

@stanley2014metaregression and @stanley2017limitations reported simulations based on a subtly different appoach to implementing the selective reporting process.
Rather than treating the selection process as stochastic, given the p-value of the generated effect size estimate, they conceptualized the process as a mixture of two types of studies (or two types of researchers).
One type of researcher, comprising $\pi_1$ of the population, reports all results regardless of whether they are affirmative.
The other type of researcher, comprising the remaining $1 - \pi_1$ of the population, only reports results that are positive and statistically significant.
In the @stanley2014metaregression simulations, every meta-analytic dataset consists of $k \times \pi_1$ effect size estimates generated by the first type of researcher and $k \times (1 - \pi_1)$ effect size estimates generated by the second type.
Both types of researchers draw primary study sample sizes from the same distribution (i.e., researcher type is independent of $\sigma_i^*$).
The observed effect size estimates therefore follow a mixture distribution in which
$$
\begin{aligned}
\Pr(T_i = t | \sigma_i) &= \pi_1 \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right) + (1 - \pi_1) \times \frac{a_i \phi\left(\frac{t - \mu}{\eta_i}\right)}{\eta_i\left[1 - \Phi(c_{1i}) \right]} \\
&= \left[\pi_1+ (1 - \pi_1) \times \frac{a_i}{\left[1 - \Phi(c_{1i}) \right]} \right] \times \frac{1}{\eta_i}\phi\left(\frac{t - \mu}{\eta_i}\right),
\end{aligned}
$$ {#eq-mixture-density}
with conditional expectation
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \frac{(1 - \pi_1) \phi(c_{1i})}{1 - \Phi(c_{1i})}.
$$ {#eq-mixture-expectation}
The question is then how the mixing proportion $\pi_1$ relates to the selection parameter $\lambda_1$.

If we wanted to directly equate the distributions for a given standard error $\sigma_i$, we could set the mixing proportion to 
$$
\pi_{1i} = \frac{\lambda_1}{1 - (1 - \lambda_1)\Phi(c_{1i})}.
$$
The difficulty here is that the mixing proportion is a function of the standard error for effect size $i$, so studies with different standard errors would need to have different mixing proportions. 
That's not how @stanley2014metaregression implement their data-generating process, so it doesn't really work as a way to translate between the selection parameter and the mixing proportion.
However, we could try to equate the unconditional oracle-weighted expectations, so that the bias of a random effects meta-analysis would be close to equal under both models.

Under the @stanley2014metaregression approach, the conditional probability of observing an effect with standard error $\sigma_i^*$ is
$$
\Pr(O_i = 1 | \sigma_i^*) = \pi_1 + (1 - \pi_1) \left[1 - \Phi(c_{1i})\right],
$$ {#eq-observation-prob-mixture}
and the unconditional expectation of an observed effect size is
$$
\mathbb{E} \left[ \frac{\mathbb{E}(T_i | \sigma_i)}{\omega \eta_i^2} \right] = \mu + (1 - \pi_1) \times \mathbb{E}\left[\frac{ \phi(c_{1i})}{\omega \eta_i \left[1 - \Phi(c_{1i})\right]} \right]
$$
Equating the unconditional oracle-weighted expectations under each model gives
$$
(1 - \lambda_1) \times \mathbb{E}\left[\frac{\phi(c_{1i})}{\omega \eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i})\right]} \right] = (1 - \pi_1) \times \mathbb{E}\left[\frac{ \phi(c_{1i})}{\omega \eta_i \left[1 - \Phi(c_{1i})\right]} \right],
$$ {#eq-equating-expectations}
which leads to
$$
\pi_1 = \lambda_1 \times \frac{\mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right] \left[1 - (1 - \lambda_1)\Phi(c_{1i})\right]} \right]}{\mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right]} \right]}.
$$ {#eq-pi1-marginally-equated}
Note that the translation from $\lambda_1$ to $\pi$ will generally depend on _all_ the parameters of the model as well as on the distribution of the observed $\sigma_i$'s.[^simpler-approx]

[^simpler-approx]: A slightly simpler approximate equating would be to take
$$
\pi_{1}' = \frac{\lambda_1}{1 - (1 - \lambda_1)\mathbb{E}\left[\Phi(c_{1i})\right]}.
$$

Computing $\pi_1$ will require some way of approximating the expectations in @eq-pi1-marginally-equated.
If $\sigma_i$ only takes on a finite set of distinct values, the expectation could be computed directly.
Otherwise, stochastic approximation could be used. 
For instance, suppose that $4 / \sigma_i^2 \sim Poisson(30)$, so we can simulate a large set of $\sigma_i$'s by taking
```{r}
#| code-fold: show

sigma_i <- 2 / sqrt(rpois(10000, lambda = 30))
```
Given values for the other model parameters, we compute the expectations by averaging over all the $\sigma_i$'s. Here is a function that computes $\pi_1$ given $\mu$, $\tau$, $\lambda_1$, $\alpha_1$, and a vector of $\sigma_i$ values:
```{r}
#| code-fold: show

mixing_prop <- function(
  mu, tau, lambda_1, sigma_i, 
  alpha_1 = .025
) {
  eta_i <- sqrt(tau^2 + sigma_i^2)
  c_1i <- (sigma_i * qnorm(1 - alpha_1) - mu) / eta_i
  d_1i <- dnorm(c_1i)
  P_1i <- pnorm(c_1i)
  num <- mean(d_1i / (eta_i * (1 - P_1i) * (1 - (1 - lambda_1) * P_1i)))
  denom <- mean(d_1i / (eta_i * (1 - P_1i)))
  pi_1 <- lambda_1 * num / denom
  return(pi_1)
}

mixing_prop(mu = 0.2, tau = 0.1, lambda_1 = 0.2, sigma_i = sigma_i)

```

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: 100%
#| fig-column: body
#| fig-cap: "Mixing proportion $\\pi_1$ as a function of selection parameter $\\lambda_1$ for selected values of $\\mu$ and $\\tau$."
#| lightbox: true
#| label: fig-mixing-prop

library(tidyverse)
param_grid <- 
  expand_grid(
    mu = seq(0.0, 0.6, 0.2),
    tau = seq(0.0, 0.4, 0.1),
    lambda_1 = seq(0, 1, 0.02)
  ) %>%
  mutate(
    mu_f = formatC(mu, format = "f", digits = 1) |> 
      fct() |> 
      fct_relabel(\(x) paste("mu ==", x)),
    tau_f = formatC(tau, format = "f", digits = 1) |> fct(),
    pi_1 = pmap_dbl(
      .l = list(mu = mu, tau = tau, lambda_1 = lambda_1),
      .f = mixing_prop, 
      sigma_i = sigma_i
    )
  )

ggplot(param_grid) + 
  aes(lambda_1, pi_1, color = tau_f) + 
  geom_line() + 
  facet_wrap(~ mu_f, labeller = "label_parsed") + 
  theme_minimal() + 
  theme(legend.position = "top") + 
  labs(
    x = expression(lambda[1]),
    y = expression(pi[1]),
    color = expression(tau)
  )
```

@fig-mixing-prop shows how the mixing proportion changes as a function of the selection parameter $\lambda_1$ for several different values of the average effect size $\mu$ and between-study heterogeneity $\tau$, using the same set of $\sigma_i$'s simulated above.
It is a fairly complicated function.
For smaller average effects, the degree of heterogeneity has relatively more influence on the shape of the function, whereas for larger average effects, the degree of heterogeneity hardly matters.
And everything depnds on the $\sigma_i$ distribution;  changing it would shift all of the curves in the picture. 

An important premise of this translation from $\lambda_1$ to $\pi_1$ is that the expectations are computed over the distribution of _observed_ $\sigma_i$'s. Because the conditional probability of observation is different under the @mcshane2016adjusting approach than under the @stanley2014metaregression approach, using a given distribution to _generate_ $\sigma_i$'s before selection will not generally lead to the same distribution of observed $\sigma_i$'s after selection.
Thus, the equating will not be as close if we hold the generating distribution constant.[^generating-distributions] 

[^generating-distributions]: One way to address this issue would be to modify the generating distribution used for each approach, so that the observed distribution of $\sigma_i$'s is held constant. This would require, for instance, sampling $\sigma_i$'s with probability proportional to $1 / \Pr(O_i = 1 | \sigma_i^*)$ under each model (i.e., @eq-observation-prob-3PSM or @eq-observation-prob-mixture).

## More steps

Paulina's simulation work also considers a more general version of a step-function selection model with steps at both $\alpha_1 = .025$ and $\alpha_2 = .500$--a four-parameter selection model (4PSM), if you will. If we let $m_i^* = I(\alpha_1 \leq p_i^* < \alpha_2)$ and $n_i^* = I(\alpha_2 \leq p_i^*)$, the selection process is
$$
\Pr(O_i = 1 | T_i^*, \sigma_i^*) = n_i^* \lambda_2 + m_i^* \lambda_1 + (1 - m_i^* - n_i^*).
$$ {#eq-selection-process-4PSM}
Following a data-generating process like the @mcshane2016adjusting approach, the probability that an effect is observed, conditional only on $\sigma_i^*$ is
$$
\Pr(O_i = 1 | \sigma_i^*) = 1 - \Phi(c_{1i}) + \lambda_1 \left[\Phi(c_{1i}) - \Phi(c_{2i}) \right] + \lambda_2 \Phi(c_{2i}).
$$ {#eq-observation-prob-4PSM}
where $c_{1i} = \left[\sigma_i \Phi^{-1}(1 - \alpha_1) - \mu\right] / \eta_i$ (as before) and $c_{2i} = \left[\sigma_i \Phi^{-1}(1 - \alpha_2) - \mu\right] / \eta_i$ and the conditional expectation of an observed effect size estimate is
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \frac{(1 - \lambda_1) \phi(c_{1i}) + (\lambda_1 - \lambda_2) \phi(c_{2i})}{1 - (1 - \lambda_1)\Phi(c_{1i}) - (\lambda_1 - \lambda_2) \Phi(c_{2i})}.
$$ {#eq-4PSM-expectation}

How would the @stanley2014metaregression approach work with more than a single step? One way to generalize it would be to think of the observed effect sizes as a three-component mixture, where there is a fraction of studies $\pi_2$ where an effect size estimate is always reported regardless of sign or statistical significance, another fraction of studies $\pi_1 > 0$ where an  effect size estimate is reported only when $p_i^* < \alpha_2$ (i.e., when the effect size estimate is larger than zero), and the remaining faction $1 - \pi_1 - \pi_2$ where an effect size estimate is reported only when $p_i^* < \alpha_1$.
Clearly, this requires that $pi_1 > 0$, $\pi_2 > 0$, and $\pi_1 + \pi_2 < 1$.
Under this data-generating process, 
$$
\Pr(O_i = 1 | \sigma_i^*) = \pi_2 + \pi_1 \left[1 - \Phi(c_{2i})\right] + (1 - \pi_1 - \pi_2) \left[1 - \Phi(c_{1i})\right],
$$ {#eq-observation-prob-mixture}
and the conditional expectation of an observed effect size estimate is 
$$
\mathbb{E}(T_i | \sigma_i) = \mu + \eta_i \times \left[\frac{\pi_1 \phi(c_{2i})}{1 - \Phi(c_{2i})} + \frac{(1 - \pi_1 - \pi_2) \phi(c_{1i})}{1 - \Phi(c_{1i})}\right].
$$ {#eq-mixture-2-expectation}
To put this approach into practice and be able to compare it to the @mcshane2016adjusting approach, we need to be able to translate from $\left(\lambda_1, \lambda_2\right)$ to $\left(\pi_1, \pi_2\right)$, at least for selection parameters where $\lambda_2 \leq \lambda_1$. 

Just as with the simpler single-step model, one way to calibrate the model parameters is to set the unconditional oracle-weighted expectations equal. 
Sufficient conditions for such equality are that
$$
\begin{aligned}
(1 - \lambda_1) \times \mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i}) - (\lambda_1 - \lambda_2) \Phi(c_{2i})\right]} \right] &= (1 - \pi_2) \times \mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right]}\right] \\
(\lambda_1 - \lambda_2)  \times \mathbb{E}\left[\frac{\phi(c_{2i})}{\eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i}) - (\lambda_1 - \lambda_2) \Phi(c_{2i})\right]} \right] &= \pi_1 \times \mathbb{E}\left[\frac{\phi(c_{2i})}{\eta_i \left[1 - \Phi(c_{2i})\right]} - \frac{\phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right]}\right]
\end{aligned}
$$ {#eq-equating-expectations-2}
or that
$$
\begin{aligned}
\pi_1 &= (\lambda_1 - \lambda_2) \times \frac{\mathbb{E}\left[\frac{\phi(c_{2i})}{\eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i}) - (\lambda_1 - \lambda_2) \Phi(c_{2i})\right]} \right]}{\mathbb{E}\left[\frac{\phi(c_{2i})}{\eta_i \left[1 - \Phi(c_{2i})\right]} - \frac{\phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right]}\right]} \\
\pi_2 &= 1 - (1 - \lambda_1) \times \frac{\mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - (1 - \lambda_1)\Phi(c_{1i}) - (\lambda_1 - \lambda_2) \Phi(c_{2i})\right]} \right]}{\mathbb{E}\left[\frac{ \phi(c_{1i})}{\eta_i \left[1 - \Phi(c_{1i})\right]}\right]}.
\end{aligned}
$$ {#eq-pi1-pi2-marginally-equated}
This particular set of conditions has the advantage that when $\lambda_2 = \lambda_1$, $\pi_1 = 0$ and $\pi_2$ then maps directly to $\lambda_1$ in the same way as under the simpler, single-step model, as given in @eq-pi1-marginally-equated.

Here is a function that computes $\left(\pi_1,\pi_2)$ given $\mu$, $\tau$, $\lambda_1$, $\lambda_2$, $\alpha_1$, $\alpha_2$, and a vector of $\sigma_i$ values:
```{r}
#| code-fold: show

mixing_prop2 <- function(
  mu, tau, lambda_1, lambda_2, sigma_i, 
  alpha_1 = .025, alpha_2 = .500
) {
  eta_i <- sqrt(tau^2 + sigma_i^2)
  
  c_1i <- (sigma_i * qnorm(1 - alpha_1) - mu) / eta_i
  d_1i <- dnorm(c_1i)
  P_1i <- pnorm(c_1i)
  num <- mean(d_1i / (eta_i * (1 - P_1i) * (1 - (1 - lambda_1) * P_1i)))
  denom <- mean(d_1i / (eta_i * (1 - P_1i)))
  pi_1 <- lambda_1 * num / denom
  
  c_i <- (tcrossprod(sigma_i, qnorm(1 - c(alpha_1,alpha_2))) - mu) / eta_i
  d_i <- dnorm(c_i)
  P_i <- pnorm(c_i)
  f_i <- eta_i * (1 - (1 - lambda_1) * P_i[,1] - (lambda_1 - lambda_2) * P_i[,2])
  g_1i <- d_i[,1] / (eta_i * (1 - P_i[,1]))
  g_2i <- d_i[,2] / (eta_i * (1 - P_i[,2]))
  pi_1 <- (lambda_1 - lambda_2) * mean(d_i[,2] / f_i) / mean(g_2i - g_1i)
  pi_2 <- 1 - (1 - lambda_1) * mean(d_i[,1] / f_i) / mean(g_1i)
  return(data.frame(pi_1 = pi_1, pi_2 = pi_2))
}

debug(mixing_prop2)
mixing_prop2(mu = 0.2, tau = 0.1, lambda_1 = 0.2, lambda_2 = 0.1, sigma_i = sigma_i)

```